A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-02 tex/body.tex
more

A rule can be applied to $\Ec$ if all the specified preconditions are met.
The conclusion of a rule either describes equalities to be added to $\Ec$
or is $\bot$ (contradiction).
% (in which case we call an application of it \emph{nonterminal})
% (in which case we call an application of it \emph{terminal}).
A rule may have multiple conclusions, % separated by $\ror$,
denoting branching.
%
An application of a rule is \emph{redundant} if at least one branch in the
conclusion is an assignment to $\Ec$ that leaves it unchanged. A
\emph{derivation tree} is a finite tree whose nodes are finite sets of
equalities and where child nodes are obtained by a nonredundant application of a
derivation rule to the parent node. A derivation tree is \emph{closed} if all of
its leaf nodes are $\bot$. A node is \emph{saturated} if no nonredundant
instance of a rule can be applied to it.

@@

Given the mapping $\Val$, the \rn{Acyclic} and \rn{Unique} rules operate as follows.
For acyclicity, if $\ec{t\typ{\delta}}$ is a datatype equivalence class
whose values $\Val \ec{t} = \MU x.\; u$ are cyclic
(as indicated by the precondition $x \in \FV( u )$),
then $\Ec$ is unsatisfiable.
For uniqueness, if $\ec{t}$ and $\ec{u}$ are two codatatype equivalence classes
whose values $\Val \ec{t}$ and $\Val \ec{u}$ are $\alpha$-equivalent,
%for a capture-avoiding renaming of $\mu$-bound variables,
then $t \teq u$. This rule may seem too restrictive, since
$\MU x.\; \const{S}(x)$ and $\MU y.\; \const{S}(\const{S}(y))$ specify the same
value despite being $\alpha$-disequivalent, but even in such cases the rule will
make some progress by discovering that the subterm $\const{S}(y)$ of $\MU y.\;
\const{S}(\const{S}(y))$ must be equal to the entire term.

@@

\paragraph{Phase 3: Branching.}
If a selector is applied to a term $t$, or if $t$'s type is a finite datatype,
the equivalence class of $t$ must contain an application of a
$\delta$ constructor.
This is enforced in the third phase by the \rn{Split} rule.
A companion rule, \rn{Single}, focuses on the degenerate case where two
terms from $\tEc$ have a corecursive singleton type
(Section~\ref{sec:the-theory-of-co-datatypes}) and must therefore be equal. Both
the finiteness assumption on the datatype in \rn{Split} and the singleton
constraint on the codatatype in \rn{Single} can be computed statically
based on a recursive computation of the cardinalities of the
constructors' argument types.

>>>>

A rule can be applied to $\Ec$ if all the specified preconditions are met.
The conclusion of a rule either describes equalities to be added to $\Ec$
or is $\bot$ (contradiction).
% (in which case we call an application of it \emph{nonterminal})
% (in which case we call an application of it \emph{terminal}).
A rule may have multiple conclusions, % separated by $\ror$,
denoting branching.
%
An application of a rule is \emph{redundant} if at least one branch in the
conclusion is an assignment to $\Ec$ that leaves it unchanged. A
\emph{derivation tree} is a %finite
tree whose nodes are finite sets of
equalities and where child nodes are obtained by a nonredundant application of a
derivation rule to the parent node. A derivation tree is \emph{closed} if all of
its leaf nodes are $\bot$. A node is \emph{saturated} if no nonredundant
instance of a rule can be applied to it.

@@

Given the mapping $\Val$, the \rn{Acyclic} and \rn{Unique} rules operate as follows.
For acyclicity, if $\ec{t\typ{\delta}}$ is a datatype equivalence class
whose values $\Val \ec{t} = \MU x.\; u$ are cyclic
(as indicated by the precondition $x \in \FV( u )$),
then $\Ec$ is unsatisfiable.
For uniqueness, if $\ec{t}$, $\ec{u}$ are two codatatype equivalence classes
whose values $\Val \ec{t}$, $\Val \ec{u}$ are $\alpha$-equivalent,
%for a capture-avoiding renaming of $\mu$-bound variables,
then $t \teq u$. This rule may seem too restrictive, since
$\MU x.\; \const{S}(x)$ and $\MU y.\; \const{S}(\const{S}(y))$ specify the same
value despite being $\alpha$-disequivalent, but the rule will
make some progress by discovering that the subterm $\const{S}(y)$ of $\MU y.\;
\const{S}(\const{S}(y))$ must be equal to the entire term.

@@

\paragraph{Phase 3: Branching.}
If a selector is applied to a term $t$, or if $t$'s type is a finite datatype,
the equivalence class of $t$ must contain a
$\delta$ constructor term.
This is enforced in the third phase by the \rn{Split} rule.
A companion rule, \rn{Single}, focuses on the degenerate case where two
terms from $\tEc$ have a corecursive singleton type
(Section~\ref{sec:the-theory-of-co-datatypes}) and must therefore be equal. Both
the finiteness assumption on the datatype in \rn{Split} and the singleton
constraint on the codatatype in \rn{Single} can be computed statically
based on a recursive computation of the cardinalities of the
constructors' argument types (and assuming all types in $\Nondata$ are
infinite).
